---
title: "The Within-Between Model"
author: "Michael Liou"
format: 
  html:
    mainfont: Palitino
    code-fold: true
    theme: lumen
    linestretch: 1.2
    fontsize: 1em
    html-math-method: katex
execute:
  cache: true
---

```{r}
#| message: false
#| code-fold: true
#| code-summary: Load Libraries

library(tidyverse)
library(nlme)
library(lme4)
library(plm)
library(kableExtra)
```

## Introduction

This material is largely sourced from Chapter 9 of Longitudinal Data Analysis.

For fixed effects models,

- $X_{ij}$ is independent of $\varepsilon_{ij}$ and $\varepsilon_{ij'}$.
  - an observation will not help determine the value of the next
- FE models "clear out" any higher level trends, and just focus on the within estimate
- The $\hat\beta$ estimate will be consistent even when the higher level effect is correlated with the covariates.

## The Model 

To introduce the model, we will look at a simulated example where there is a higher level effect

```{r}
#| code-fold: true

# simulated  
dat_skel <- data.frame(cohort = rep(1:3, each = 16),
           id = rep(1:12, each = 4),
           age = c(rep(c(4, 6, 7, 10), 4), # cohort 1
                   rep(c(4, 5, 8, 9) + 1, 4), # cohort 2 
                   rep(c(4, 7, 8, 10) + 2, 4))) # cohort 3
# creating data
set.seed(1)
id_int <- rnorm(12, 2) + 10
id_slope <- rnorm(12, mean = 1, sd = .01) # random slope
cohort_int <- c(1, 4, 7)

dat <- dat_skel %>% 
  mutate(y = id_int + id_slope * age + cohort_int[cohort] + rnorm(nrow(.), sd = 5),
         id = factor(id),
         cohort = factor(cohort))

dat %>% head()
```
 
```{r}
#| code-fold: true
dat %>% ggplot(aes(age, y, color = cohort, group = id)) +
  geom_point() +
  geom_line()
```

```{r}
dat_centered <- dat %>% arrange(id, age) %>%
  group_by(id) %>% 
  mutate(age_mean = mean(age),
         y_mean = mean(y),
         age_centered = age - age_mean,
         y_centered = y - y_mean,
         age_baseline = age - first(age),
         y_baseline = y - first(y))

dat_baseline <- dat_centered %>% 
  filter(age_baseline != 0)
```


```{r}
# Fixed Effect Estimate (within model)
centered_lm <- lm(y_centered ~ age_centered, data = dat_centered) # centered regression

# between model
baseline_lm <- lm(y_baseline~age_baseline, data = dat_baseline) # baseline adjusted (centered)
id_lm <- lm(y~age + id, data = dat_centered) # id model (centered)

mean_lm <- lm(y_mean ~ age_mean, data = dat_centered) # mean regression (pooled model)

pooled_lm <- lm(y~age, data = dat_centered)

cohort_lm <- lm(y~age + cohort, data = dat_centered)

# mixed model
mmod <- lmer(y ~ age + (1|id), data = dat)

# within-between model
wb_mmod <- lmer(y~age_centered + age_mean + (1 | id), data = dat_centered)
```


```{r}
#| code-fold: true
cbind(
  baseline_lm = coef(baseline_lm),
  mean = coef(mean_lm),
  cohort = coef(cohort_lm)[1:2],
  id = coef(id_lm)[1:2],
  pooled = coef(pooled_lm),
  centered = coef(centered_lm),
  mixed = fixef(mmod)) %>% rbind(NA) %>% 
  cbind(within_between = fixef(wb_mmod))
```

$$
\begin{aligned}
w = \frac{(1 - \rho_y)\rho_x}{(1 - \rho_y) + n\rho_y(1-\rho_x)}
\end{aligned}
$$
and 

$$
\begin{aligned}
\hat \beta_2^{(RE)} = (1 - w)\hat\beta_2^{(FE)} + w\hat\beta_2^{(B)}
\end{aligned}
$$

```{r}
# help function for weight
fixed_random_weight <- function(rho_x, rho_y, n) {
  (1-rho_y) * rho_x / ((1-rho_y) + n * rho_y * (1 - rho_x))
}

```

```{r}
# response correlation
rho_y <- VarCorr(mmod) %>% as.data.frame() %>% pull(vcov) %>% 
  as_mapper(~.x[1]/sum(.x))()

# correlation of age
age_lm <- lm(age~id, data = dat_centered)
rho_x <- summary(age_lm)$r.squared

n <- 4
w <- fixed_random_weight(rho_x, rho_y, n)
(1-w) * fixef(wb_mmod)[2] + w * fixef(wb_mmod)[3]
```



```{r}
dat %>% ggplot(aes(age, y, color = cohort, group = id)) +
  geom_point(alpha = .2) +
  geom_line(alpha = .2) +
  geom_abline(slope = coef(centered_lm)[2], intercept = coef(centered_lm)[1] + 10) + # centered estimate
  geom_abline(slope = coef(mean_lm)[2], intercept = coef(mean_lm)[1]) + # mean estimate
  geom_abline(slope = fixef(mmod)[2], intercept = fixef(mmod)[1], linetype = 2) + # mixed effects estimate
  theme_minimal()
```





## FEV example


```{r}
library(foreign)
fev <- read.dta("data/fev1.dta") %>% 
  filter(id != 197) %>%
  mutate(id = factor(id),
         y =  logfev1 - 2 * (log(ht)))

fev_centered <- fev %>% group_by(id) %>% 
  mutate(
    y_mean = mean(y),
    age_mean = mean(age),
    age_centered = age - age_mean)
```



```{r}
# fitting all models
fev_lm <- lm(y ~ age + id, data = fev) # lm
fev_mer <- lmer(y~age + (1|id), data = fev) # random

# within/between model
fev_wb <- lmer(y ~ age_centered + age_mean + (1|id),
               data = fev_centered,
               REML = FALSE)

# cross sectional
fev_mean_lm <- lm(y_mean ~ age_mean, data = fev_centered)


# age correlation (between/within)
fev_age_lm <- lm(age ~ id, data = fev)

```

```{r}
# comparing the coefficients
cbind(fixed = coef(fev_lm)[1:2],
      random = fixef(fev_mer),
      mean = coef(fev_mean_lm)) %>% 
  rbind(NA) %>% cbind(within_between = fixef(fev_wb))
```



```{r}
rho_y <- VarCorr(fev_mer) %>% 
  as.data.frame() %>%
  pull(vcov) %>%
  as_mapper(~.x[1]/sum(.x))() # .689

# age covariance
# fev_age_ss <- anova(fev_age_lm)$`Sum Sq`
# rho_x <- fev_age_ss[1] / sum(fev_age_ss) 
rho_x <- summary(fev_age_lm)$r.squared # .172

nbar <- fev %>% count(id) %>% pull(n) %>% mean()

w <- fixed_random_weight(rho_x, rho_y, nbar)
w
```

Since this value of w is very low we expect that the mixed estimate is very close to the fixed estimate.


```{r}
# hausman test manually

haus_contr <- cbind(c(0, -1, 1))
haus_mean <- fixef(fev_wb) %*% haus_contr
haus_var <- t(haus_contr) %*% vcov(fev_wb) %*% haus_contr
haus_z <- haus_mean / sqrt(haus_var)[1]
pchisq(haus_z^2, 1, lower.tail = FALSE) # p = .84

# hausman with plm

# phtest(y~age, data = fev_centered, index = c("id"))
wi <- plm(y~age, data = fev_centered, index = c("id"), model = "within" )
fe <- plm(y~age, data = fev_centered, index = c("id"), model = "between")
mi <- plm(y~age, data = fev_centered, index = c("id"), model = "random")

phtest(wi, mi)
```

## Resources

- [Separation of individual-level and cluster-level covariate effects in regression analysis of correlated data](https://onlinelibrary-wiley-com.ezproxy.library.wisc.edu/doi/pdfdirect/10.1002/sim.1524)
