---
title: "Time Series"
author: "Michael Liou"
date: "`r Sys.Date()`"
execute:
  cache: true
---


```{r setup, include=FALSE}
#| code-summary: Libraries
#| message: false
library(tidyverse) # general data science
library(glue)
library(reshape2) # for melt function
library(astsa)
library(tseries)
library(forecast) # time series forecasting, common package
library(vars) # vector autoregression
library(marima) # multivariate arima
library(dlm) # dynamic linear models
library(ggfortify) # adds some autoplot models, overwrites many functions from forecast
library(KFAS) # kalman filter, sequential univariate (fast), exponential family
# detach("package:forecast", unload = T)
```

# Introduction

## Types of Models

- Autoregression (AR)
- Moving Average (AM)
- Vector Autoregression (VAR)

- HMM
  - related to bayesian inference, special
- State Space Models
  - very broad space of models that includes arima models, and dynamic linear models. Can also account for linear/nonlinear gaussian/non-gaussian errors
- Multivariate autoregression (MAR)

## Software

- `stats`
  - `KalmanLike`, `KalmanRun`, `KalmanSmooth`, `KalmanForecast`
- `dlm`
- `KFAS`

See also [State Space Models in R](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwjng_vXx4P6AhWdk4kEHZ4KBUAQFnoECBoQAQ&url=https%3A%2F%2Fwww.jstatsoft.org%2Farticle%2Fdownload%2Fv041i04%2F488&usg=AOvVaw0bbJoxE4fWu74zhZs0qggz).

## Theory

Courses 

- Kevin Kotzé's [Time Series Analysis Course](https://www.economodel.com/time-series-analysis)
  - has a great overview and description, and packaged R commands for univariate and multivariate from an economics perspective.

# Univariate time series

## base tools

```{r}
# acf, ccf, pacf
```

## AR(1)

Autoregressive of order 1 is the simplest time series you can have.

$$
\begin{aligned}
Y_t = Y_{t-1} + \varepsilon_t
\end{aligned}
$$

## Simulated Examples

```{r}
# AR models
sim_ar <- tibble(p = 1:3,
                 ar = list(.7,
                           c(.7,-.3),
                           c(.7, -.3, .5)),
                 n = 200) %>% 
  rowwise() %>% 
  mutate(y = list(arima.sim(list(ar = ar), n = 200)),
         x = list(time(y)))

sim_ar %>% unnest(c(x, y)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  facet_grid(p~.)
```


```{r}
# ma processes
sim_ma <- tibble(q = 1:3,
                 ma = list(.7,
                           c(.7,-.3),
                           c(.7, -.3, .5)),
                 n = 200) %>% 
  rowwise() %>% 
  mutate(y = list(arima.sim(list(ma = ma), n = 200)),
         x = list(time(y))) %>% 
  ungroup()

sim_ma %>% unnest(c(x, y)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  facet_grid(q~.)
```





## Examples: LA County

We just examine the cardiovascular mortality from the LA Pollution study. These are average weekly cv mortality rates from `astsa` package.

```{r}
# ?cmort
x <- cmort %>% as.numeric()
x1 <- x %>% lag()
# yt = u + beta * yt-1
cmort_lm <- lm(x~x1)

cbind(coef(cmort_lm)[1] + coef(cmort_lm)[2] * x1,
      c(NA, fitted(cmort_lm)),
      c(NA, predict(cmort_lm))) %>% 
  head() # fitted/predicted values match up


```


```{r}
qplot(time(cmort), x, geom = "line", color = "black") +
  geom_line(aes(y =  c(NA, fitted(cmort_lm)), color = "red")) +
  scale_color_manual(values = c("black", "red"),
                     label = c("raw", "fitted"))
```



## Example: AirPassengers

```{r}
AirPassengers %>% plot()
```

```{r}
sarima(AirPassengers, d = 1, p = 2, q = 0)
```


# Multivariate time series

## Vector Autoregression

### Example: LA County (VAR)


```{r}
# cardiovascular mortality, temperature and particulates in LA county, weekly.
la <- cbind(cmort, tempr, part)

# visualization
# ts.plot(cmort, tempr, part, col = 1:3) # base r of autoplot 
autoplot(la, color = "black") + facet_grid(series~., scales = "free_y")
```


```{r}
la_var1 <- VAR(la, p = 1, type = "both") # w/ trend
la_var2 <- VAR(la, p = 2, type = "both") # w/ trend

# coefficient matrix
sapply(coef(la_var1),rlang::as_function(~.x[,"Estimate"]))

# vcov of coef estimates
sapply(coef(la_var1), rlang::as_function(~round(.x[,"Std. Error"], 3)))

# estimated covariance of errors
summary(la_var1)$covres
```


```{r}
la_var1_const <- VAR(la, p = 1, type = "const")

sapply(coef(la_var1_const),rlang::as_function(~.x[,"Estimate"]))
```


```{r}
# 
la_ols <- ar.ols(la, order.max = 1, demean = FALSE, intercept = TRUE)

# matches constant VAR coefficients
rbind(t(la_ols$ar[1,,]),
      const = la_ols$x.intercept)
      
```


```{r}
la %>% melt(c("time", "series")) %>% 
  ggplot(aes(time, value)) + 
  geom_line(color = "black") + 
  geom_line(data = fitted(la_var1) %>% melt(c("time", "series")),
            mapping = aes(time, value, color = "red"), alpha= .6) +  # var1
  geom_line(data = fitted(la_var1_const) %>% melt(c("time", "series")),
            mapping = aes(time, value, color = "blue"), alpha = .6) + # var1 const
  geom_line(data = fitted(la_var2) %>% melt(c("time", "series")),
            mapping = aes(time, value, color = "green"), alpha = .6) + # var2
  scale_color_manual(values = c("red", "blue", "green"),
                     labels = c("var1", "var1_const", "var2")) + 
  facet_grid(series~.)
```

From looking at the models, they're quite difficult to tell which model is fitting better than the others.



```{r}
# var estimates using lm
# Matrix of time series
la_mat <- la %>% `class<-`("matrix")

la_var1_manual <- lm(la_mat ~ cbind(lag(la_mat), 1:nrow(la_mat))) # w/ trend
la_var1_const_manual <- lm(la_mat ~ lag(la_mat)) # const
```

```{r}
# coefs match
sapply(coef(la_var1),rlang::as_function(~.x[,"Estimate"]))
coef(la_var1_manual)[c(2:4, 1, 5),] %>% # reorder to match
  `rownames<-`(c("cmort", "tempr", "part", "const", "trend")) 

# var matches
# SSE / n-r-p
# nobs - coefs_estimated - VAR_order
crossprod(la_var1_const_manual$residuals) / (nrow(la_mat) - la_var1_const_manual$rank - 1) # 503
summary(la_var1_const)$covres
```

```{r}
# order selection
VARselect(la) # selects 2 by BIC
```


```{r}
acf(resid(la_var2), 52)$acf
```

The CCF plots should all be non significant. The second part of "x & y" are the ones that lead.

```{r}
# serial test
serial.test(la_var2, lags.pt = 12, type = "PT.adjusted")
la[,1]
```

There are some large sample properties of VAR's as well,

```{r}
library(marima)
```


# State Space Model

These models have a hierarchical form, in which there is some underlying time series process, but then we also observe data on top of that. Thus, the general form of the equations look like this:

$$
\begin{aligned}
\textbf{State Equation:}& \\
x_t &= \Phi x_{t-1} + \Upsilon u_t +   w_t \\
\textbf{Observation Equation:}& \\
y_t &= A_t x_t + \Gamma u_t  + v_t
\end{aligned}
$$
where:

- $\Upsilon u_t$ - is time varying exogenous variables
- $\Upsilon u_t$ - is time varying exogenous variables
- $x_t$ is state at time $t$
- $\Phi x_t$ is state at time $t$ describes how $x$ evolves
- $w_t$ is state noise with $w_t \sim N(0, Q)$
- $v_t$ is measurement noise with $v_t \sim N(0, R)$

There are more general forms of the state space model: with correlated errors, see Durbin Koopman for more state space methods.

It seems now that state space models are now also being superseded by recurrent neural networks, which can model dynamical properties.

## Example: AR(1) with observational noise

The state equation:

$$
\begin{aligned}
\textbf{State Equation:}& \\
x_t &= \phi x_{t-1} + w_t \\
\textbf{Observation Equation:}& \\
y_t &= x_t + v_t
\end{aligned}
$$

What is interesting about this case with a hierarchical data structure, is that we can show it has the same error structure as an ARMA model. Shumway Stoffer notes that even though it has the same parameterization as an ARMA model, it is often easier to think about the state model form. See example 6.3 for more details.

## Kalman Filter

Kalman filter is a recursive, markovian updating algorithm for estimating a hidden state variable given noisy and partial observations. The common example is that we are tracking a truck by gps observations. Since the gps observations are imprecise, they will jump back and forth.

An excellent resource explanation with pictures is found [from bzarg](https://www.bzarg.com/p/how-a-kalman-filter-works-in-pictures/)

### Example: Local level Model

Consider the equations:

$$
\begin{aligned}
\textbf{State Equation:}& \\
x_t &= x_{t-1} + w_t \\
\textbf{Observation Equation:}& \\
y_t &= x_t + v_t
\end{aligned}
$$
where both $w_t, v_t \sim N(0, 1)$. We can make the Kalman filter here based on our observation equation. We assume that $A = 1$ (in this case) and $Phi = 1$ are known here. In reality, we can use maximum likelihood of the _innnovations_ (prediction errors) in order to estimate the 

```{r}
# generate the data
set.seed(1)
n <-  50
x0 <- rnorm(1) # random initial state
w <- rnorm(n, 0, 1) # state level noise
v = rnorm(n, 0, 1) # observation level noise
x <- cumsum(c(x0, w)) # true state variables
y <- x[-1] + v # remove initial state
ks <- Ksmooth0(num = 50, 
               y = y, 
               A = 1, 
               mu0 = 10, # set initial values
               Sigma0 = 20, # set initial values
               Phi = 1,
               cQ = 1,
               cR = 1)
```


```{r}
# plot all the predictions
par(mfrow = c(3, 1),
    mar = c(2, 4, 2, 4))
# predictions
plot(x[-1], main = "Prediction", ylim = c(-5, 10))
lines(ks$xp[1,,])
lines(ks$xp + 2*sqrt(ks$Pp[1,,]), lty = 2, col = 4)
lines(ks$xp - 2*sqrt(ks$Pp[1,,]), lty = 2, col = 4) # variance matrices

# filters
plot(x[-1], main = "Filters", ylim = c(-5, 10), xlab = "")
lines(ks$xf[1,,])
lines(ks$xf + 2*sqrt(ks$Pf[1,,]), lty = 2, col = 4)
lines(ks$xf - 2*sqrt(ks$Pf[1,,]), lty = 2, col = 4) # variance matrices


# ks$xs[1,,] # smooth values
plot(x[-1], main = "Smooth", ylim = c(-5, 10), xlab = "")
lines(ks$xs[1,,])
lines(ks$xs + 2*sqrt(ks$Ps[1,,]), lty = 2, col = 4)
lines(ks$xs - 2*sqrt(ks$Ps[1,,]), lty = 2, col = 4) # variance matrices
```

```{r}
# creating table for comparison
tibble(predict = ks$xp[1,,],
       filter = ks$xf[1,,],
       smooth = ks$xs[1,,],
       predict_sd = sqrt(ks$Pp[1,,]),
       filter_sd = sqrt(ks$Pf[1,,]),
       smooth_sd = sqrt(ks$Ps[1,,])) %>%
  mutate(across(predict:smooth, ~scales::number(.x, accuracy = .001)),
         across(predict_sd:smooth_sd, ~scales::number(.x,accuracy = .01))) %>% 
  transmute(predict = glue("{predict} ({predict_sd})"),
         filter = glue("{filter} ({filter_sd})"),
         smooth = glue("{smooth} ({smooth_sd})"))
```


## Estimation of State Parameters

There are bayesian methods for estimating the state parameters but also maximum likelihood w/ Newton Raphson or EM algorithmss

## Example: Nile {.tabset}

Annual flow of the river Nile from 1871 - 1970. There's an apparent changepoint near 1898.

I found this example going through all the state space model libraries. The orginal paper is called "JSS Journal of Statistical Software, State Space Models in R".Specifically, we walk through the Nile example with 3 functions: 

1. `stats::StructTS`
2. `dlm:dlmMLE`
3. `KFAS:kf`

```{r}
# change point analyses
ts.plot(Nile)
```

### stats::StructTS


```{r}
nile_sts <- StructTS(Nile, "level")
# nile_sts %>% tsdiag() # diagnostics of structural model

# values from model.
tibble(
  times = time(Nile),
  filtered = fitted(nile_sts)[,"level"], # filtered values
  smoothed = tsSmooth(nile_sts)[,"level"]) # smoothed valuees

# forecast values
predict(nile_sts,n.ahead = 5)

# plotting forecast values, with package forecast
plot(forecast::forecast(nile_sts, # StructST object
                   level = c(50, 90), # CI levels
                   h = 10), # periods of forecasting
     xlim = c(1950, 1980))
```

### dlm::dlm

```{r}
# set up the dlm object
nile_dlm_ll <- function(theta){
  dlmModPoly(order = 1, dV = theta[1], dW = theta[2]) # fits local level model
}

# calls optim internally to optimize model (default BFGS)
# could use numDeriv::hessian for numerically accurate evaluation of Hessians
nile_dlm_mle <- dlmMLE(Nile, # data
                   parm = c(100, 2), # initial parameters
                   nile_dlm_ll, # model
                   lower = rep(1e-4, 2)) 

nile_dlm_mle$par # similar variance parameters of local linear model
```


```{r}
nile_dlm_ll_best <- nile_dlm_ll(nile_dlm_mle$par) # build model with best fit by optim
W(nile_dlm_ll_best) # state randomnesss
V(nile_dlm_ll_best) # observation error variance

# can use fitted model to create smooth estimates now
# $s has time series of smooth estimates
# $U.S and $D.S have SVD of smoothing varriances (for std err)
nile_dlm_ll_smooth <- dlmSmooth(Nile , nile_dlm_ll_best)
# conf ints can be calculated by
```


```{r}
# calculate standard errors
hwidth <- sqrt(unlist(dlmSvd2var(nile_dlm_ll_smooth$U.S, nile_dlm_ll_smooth$D.S))) * qnorm(0.025, lower = FALSE)
nile_dlm_ll_smooth_ci <- cbind(nile_dlm_ll_smooth$s, as.vector(nile_dlm_ll_smooth$s) + hwidth %o% c(-1, 1))
autoplot(nile_dlm_ll_smooth_ci) + theme(legend.position = "") +
  labs(title = "smoothed kalman with CI") +
  geom_point(data = tibble(time = time(Nile),
                           flow = Nile,
                           series = "real"),
             mapping = aes(time, flow))
```

### KFAS::KFS

Kalman filtering/smoothing/simulation for linear state space models in the exponential family. KFAS uses the sequential processing method. This package uses a slightly different parameterization of their state space models.

$$
\begin{aligned}
y_t &= Z_t\alpha_t + \varepsilon_t &\text{observation}\\
\alpha_{t+1} &= T_t \alpha_t + R_t\eta_t &\text{transition}
\end{aligned}
$$

```{r}
# build the local linear model
# can initialize a specific model with values
# nile_kfas_ll_model <- SSModel(Nile ~ SSMtrend(1, Q = 15000), # Q = Transition error
#                               H = 30) # Observation error

# add NA for values you want to optimize
nile_kfas_ll_model <- SSModel(Nile ~ SSMtrend(1, Q = list(matrix(NA))), # Q = Transition error
                              H = matrix(NA)) # Observation error


# fit the model with wrapper to `optim`
# -logLik(nile_kfas_ll_model) # is the objective that is optimized
nile_kfas_ll_fit <- fitSSM(nile_kfas_ll_model,
       c(log(var(Nile)), log(var(Nile))), # initial parameters for optim
       method = "BFGS")

# extract just the optimal model
nile_kfas_ll <- nile_kfas_ll_fit$model

# Filter/smooth
nile_kfas_ll_smooth <- KFS(nile_kfas_ll,
    filtering = "state",
    smoothing = "state")

# autoplot calls fortify, grabbing some components of the model
# ggfortify:::autoplot.KFS
# fortify(nile_kfas_ll_smooth)  # create df with: time, raw y, alphahat (smoothed values), raw residual = raw y - alphahat
# nile_kfas_ll_smooth$alphahat # contains smoothed state variable estimates. for some reason "fitted" doesn't return these.
autoplot(nile_kfas_ll_smooth)
```


```{r}
# these give same values
# cbind(predict(nile_kfas_ll),
      # nile_kfas_ll_smooth$alphahat)

nile_kfas_ll_ci <- predict(nile_kfas_ll, interval = "confidence", level = .9)
nile_kfas_ll_pi <- predict(nile_kfas_ll, interval = "prediction", level = .9)


legend_labels <- c("lower_ci", "upper_ci", "fit", "lower_pi", "upper_pi")
# forecast:::autoplot.mts
autoplot(cbind(nile_kfas_ll_ci[,-1], nile_kfas_ll_pi),
         mapping = aes(x, y, group = series, linetype = series)) +
  scale_linetype_manual(values = c(2, 2, 1, 3, 3), labels = legend_labels) +
  scale_color_manual(values = c(2, 2, 1, 3, 3), labels = legend_labels) +
  labs(y = "Predicted Annual flow", main = "River Nile") 
```


```{r}
# forecasting
nile_kfs_mean <- KFS(nile_kfas_ll,
    filtering = c('state')) # specifying "mean" only estimates the smooths (alphahat)

cbind(predict = nile_kfs_mean$a, # one step ahead prediction
      filter = nile_kfs_mean$att, # filter estimates
      smooth = nile_kfs_mean$alphahat) %>%  # smoothed
  autoplot() + 
  coord_cartesian(ylim = c(700, 1250)) +
  labs(title = "predict/filter/smooth estiamtes of Nile with KFS")
```

