---
title: "Time Series"
author: "Michael Liou"
date: "`r Sys.Date()`"
execute:
  cache: true
---


```{r}
#| code-summary: Libraries and Setup
#| message: false
library(tidyverse)  # general data science
library(glue)
library(reshape2)   # for melt function
library(astsa)
library(kableExtra) # kableExtra
library(tseries)
library(forecast)   # time series forecasting, common package
library(vars)       # vector autoregression
library(broom)      # tidy results
library(sparsevar)  # sparse vector autoregression
library(marima)     # multivariate arima
library(dlm)        # dynamic linear models
library(ggfortify)  # adds some autoplot models, overwrites many functions from forecast
library(KFAS)       # kalman filter, sequential univariate (fast), exponential family
library(fGarch)    # garch models
library(kableExtra) # Kable extra
library(TSrepr)   # lower dimensional representation of time series
library(here) # for relative file finding


# detach("package:forecast", unload = T)

source(here("scripts/knit_latex.R"))
```

## Introduction

### Types of Models

- Autoregression (AR)
- Moving Average (AM)
- Vector Autoregression (VAR)

- HMM
  - related to bayesian inference, special
- State Space Models
  - very broad space of models that includes arima models, and dynamic linear models. Can also account for linear/nonlinear gaussian/non-gaussian errors
- Multivariate autoregression (MAR)

### Software

- `stats`
  - `KalmanLike`, `KalmanRun`, `KalmanSmooth`, `KalmanForecast`
- `dlm`
- `KFAS`

See also [State Space Models in R](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwjng_vXx4P6AhWdk4kEHZ4KBUAQFnoECBoQAQ&url=https%3A%2F%2Fwww.jstatsoft.org%2Farticle%2Fdownload%2Fv041i04%2F488&usg=AOvVaw0bbJoxE4fWu74zhZs0qggz).

### Theory

Courses 

- Kevin Kotzé's [Time Series Analysis Course](https://www.economodel.com/time-series-analysis)
  - has a great overview and description, and packaged R commands for univariate and multivariate from an economics perspective.

## ARIMAX

The general pattern for univariate time series analysis is as follows

1. figure out how to make stationary
  - by trend modeling (exogenous), or differencing
  - ["Breusch-Godfrey"](https://stats.stackexchange.com/questions/518737/durbin-vs-breusch-godfrey-test-for-autocorrelation-which-is-better) : up to p order ar
  - "Durbin-Watson" : first order ar testing
2. 

### base tools

```{r}
# acf, ccf, pacf, arima, arima.sim
```

### AR(1) and MA(1)

Autoregressive of order 1 is the simplest time series you can have.

$$
\begin{aligned}
Y_t = \alpha Y_{t-1} + \varepsilon_t
\end{aligned}
$$


### Simulated Examples

this section we'll simulate a number of AR and MA models so that you get a better sense of what they look like.



```{r}
#| fig-cap: AR Model Simulations
#| cap-location: margin
# AR models
sim_ar <- tibble(p = 1:3,
                 ar = list(.7,
                           c(.7,-.3),
                           c(.7, -.3, .5)),
                 n = 200) %>% 
  rowwise() %>% 
  mutate(y = list(arima.sim(list(ar = ar), n = 200)),
         x = list(time(y)))

sim_ar %>% unnest(c(x, y)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  facet_grid(p~.)
```


```{r}
#| fig-cap: MA Model Simulations
# MA processes
sim_ma <- tibble(q = 1:3,
                 ma = list(.7,
                           c(.7,-.3),
                           c(.7, -.3, .5)),
                 n = 200) %>% 
  rowwise() %>% 
  mutate(y = list(arima.sim(list(ma = ma), n = 200)),
         x = list(time(y))) %>% 
  ungroup()

sim_ma %>% unnest(c(x, y)) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  facet_grid(q~.)
```

### Analyzing simulations

```{r}
#| message: false
#| results: hold
# ar1
set.seed(1)
ar1 <- arima.sim(list(order = c(1, 0, 0), ar = .2), 100)

invisible(capture.output(
  mod_sar1 <- sarima(ar1, p = 1, d = 0, q = 0) # asta
  ))
mod_Ar1 <- Arima(ar1, order = c(1, 0, 0)) # forecast
mod_ar1 <- arima(ar1, order = c(1, 0, 0)) # base

bind_rows(
  tidy(mod_Ar1) |> add_column(model = "forecast::Arima", .before = 1),
  tidy(mod_ar1) |> add_column(model = "stats::arima", .before = 1),
  mod_sar1$ttable[, 1:2] |> 
  as_tibble(rownames = "term", .name_repair = ~c("estimate", "std.error")) |>
  add_column(model = "astsa::sarima", .before = 1)
) |> arrange(desc(term), model) |> 
  kable() |> kable_minimal()
```

```{r}
#| message: false
#| results: hold
# arx1
set.seed(1)
x <- sort(runif(400))
beta <- 2
arx1 <- arima.sim(list(ar=.3), 400) + x * beta

mod_Arx1 <- Arima(arx1, order = c(1, 0, 0), xreg = x)
mod_arx1 <- arima(arx1, order = c(1, 0, 0), xreg = x)
invisible(capture.output(
  mod_sarx1 <- sarima(arx1, p = 1, d = 0, q = 0, xreg = x)
))

bind_rows(
  mod_Arx1 |> tidy()|> add_column(model = "forecast::Arima", .before = 1),
  mod_arx1 |> tidy() |> add_column(model = "stats::arima", .before = 1),
  mod_sarx1$ttable[,1:2] |>
    as_tibble(rownames = "term", .name_repair = ~c("estimate", "std.error")) |>
    add_column(model = "astsa::sarima", .before = 1)
) |> arrange(desc(term)) |> 
  kbl() |> kable_minimal()
```


It's pretty clear from both of these tests, of the AR1 model and the AR1X model that they estimate very similarly,

### Diagnostics

There are hypothesis tests, and graphical checks for stationarity

unit root tests are for stability/stationarity:

- Augmented Dickey Fulley Test
  - want low p value
- KPSS Test
  - want high p value for stationarity


### Examples: LA County

We just examine the cardiovascular mortality from the LA Pollution study. These are average weekly cv mortality rates from `astsa` package.

```{r}
# ?cmort
x <- cmort %>% as.numeric()
x1 <- x %>% lag()
# yt = u + beta * yt-1
cmort_lm <- lm(x~x1)

cbind(coef(cmort_lm)[1] + coef(cmort_lm)[2] * x1,
      c(NA, fitted(cmort_lm)),
      c(NA, predict(cmort_lm))) %>% 
  head() # fitted/predicted values match up


```


```{r}
qplot(time(cmort), x, geom = "line", color = "black") +
  geom_line(aes(y =  c(NA, fitted(cmort_lm)), color = "red")) +
  scale_color_manual(values = c("black", "red"),
                     label = c("raw", "fitted"))
```


### Example: AirPassengers

```{r}
AirPassengers %>% plot()
```

```{r}
sarima(AirPassengers, d = 1, p = 2, q = 0)
```

## Exponential smoothing

- `ets()` is probably the function you want

```{r}
data(oil)
fc <- ses(oil, h = 5, alpha = .1) # can leave out the alpha to estimate alpha instead

# oil |> autoplot() +
#   geom_line(data = data.frame(time = index(fitted(fc)),
#                        fitted = fitted(fc)),
#             mapping = aes(time, fitted),
#             color = "red")

autoplot(fc) + 
  autolayer(fitted(fc)) +
  theme(legend.position ="none")

```


## TSrepr

This R package is designed to give smaller dimension representations of time series. If there are $n$ observations of the time series, this package would like to represent that as $p$. parameters. This roughly equates to nonparametric regression of the signals. We'll investigate how these work by following along in the [tutorial given](https://petolau.github.io/TSrepr-time-series-representations/):

```{r}
data("elec_load")

elec <- as.numeric(elec_load[1,])
 
ggplot(data.frame(Time = 1:length(elec), Value = elec), aes(Time, Value)) +
  geom_line() +
  theme_bw()

```


```{r}
# DWT with level of 2^3
data_dwt <- repr_dwt(elec, level = 3) # discrete wavelet transform
# first 84 DFT coefficients are extracted and then inverted
data_dft <- repr_dft(elec, coef = 84)
# first 84 DCT coefficients are extracted and then inverted
data_dct <- repr_dct(elec, coef = 84)
# Classical PAA
data_paa <- repr_paa(elec, q = 8, func = mean)
```

```{r}
data_plot <- data.frame(Value = c(data_dwt, data_dft, data_dct, data_paa),
                        Time = rep(1:length(data_dwt), 4),
                        Method = factor(rep(c("DWT", "DFT", "DCT", "PAA"),
                                            each = length(data_dwt))))
 
ggplot(data_plot, aes(Time, Value)) +
  geom_line(aes(color = Method), alpha = 0.80, size = 0.8) +
  geom_line(mapping = aes(Time, Value), data = data.frame(Time = 1:length(elec), Value = elec)) + 
  theme_bw()
```


## ARCH

Arch models have variance that depend on the previous timepoint as well. That is, an ARCH(1) model has dependence on the previous error term.

$$
\begin{aligned}
y_t &= b'x_t + \varepsilon_t \\
\varepsilon_t &\overset{i.i.d.}{\sim} N(0, \sigma^2_t) \\
\text{Var}(\varepsilon_t) = \sigma^2_t &= \omega + \alpha_1\varepsilon_{t-1}^2
\end{aligned}
$$

We can write the zero centered process a little easier, 

$$
\begin{aligned}
y_t &= \sigma_t\epsilon_t \\
\sigma_t^2 &= \omega + \alpha_1y^2_{t-1} \\
\varepsilon_t &\overset{i.i.d}{\sim} N(0, 1)
\end{aligned}
$$

Let's simulate the ARCH model and see the basic characteristics, we're simulating a zero mean 

$$
\begin{aligned}
\sigma^2_t = 5 + .5 y^2_{t-1}
\end{aligned}
$$

```{r}
# helper functions to simulate arch/garch models
sim_arch <- function(y0 = 2, omega = 5, alpha1 = .5, n = 100) {
  y <- vector("numeric", n)
  y[1] <- y0
  inn <- rnorm(n)
  for (i in 2:n) {
    y[i] <- inn[i] * sqrt(omega + alpha1 * y[i-1]^2) # arch
  }
  y
}
```

```{r}
#| layout: [[100], [50, 50], [50, 50]]
set.seed(1)
y <- sim_arch(n=1000)
plot(y, type = "l", main = "Time Series Plot")
Acf(y, main = "ACF y")
Pacf(y, main = "PACF y")
Acf(y^2, main = "ACF y^2")
Pacf(y^2, main = "PACF y^2")
```

In the time series plot, we can see why this is stochastically volatile plot, since there are regions of the time series that go way more wild when the values start to get high. The characteristics of an ARCH(1) model, are that the series $y$ look like white noise, and $y^2$ look like an AR(1) process. There's a spike in the PACF model for lag 1, and a geometric decay in the acf plot of $y^2$ indicating AR(1).

```{r}
#| message: false
#| warning: false
#| results: hold
invisible(capture.output(
  mod_arch <- garch(y, order = c(0, 1))
))
summary(mod_arch)
```

The above is the output for fitting an ARCH(1) model, and we can see that the estimates are quite close to one another for the parameters that are controlling the variance of the function. Also notice that we have almost 1000 values that we had to feed in too though, it seems that estimating the variance coefficients is quite inefficient.

## GARCH

A garch model will use the last error value, as well as the last variance for the error model.

$$
\begin{aligned}
y_t &= b'x_t + \varepsilon_t \\
\varepsilon_t &\overset{i.i.d.}{\sim} N(0, \sigma^2_t) \\
\text{Var}(\varepsilon_t) = \sigma^2_t &= \omega + \alpha_1\varepsilon_{t-1}^2 + \beta_1\sigma_{t-1}^2
\end{aligned}
$$

The following we are simulating

$$
\begin{aligned}
Var(\varepsilon_t) = \sigma^2 = .1 + .1\varepsilon_{t-1}^2 + .8 \sigma^2_{t-1}
\end{aligned}
$$

```{r}
# see also fGarch::garchSim
sim_garch <- function(y0 = 2, omega = .1, alpha1 = .1, beta1 = .8, n = 100, xt = 0, b = .4) {
  if (xt != 0 & length(xt) != n) rlang::abort("length of xt should be length of simulation")
  y <- vector("numeric", n)
  err <- vector("numeric", n)
  sds <- vector("numeric", n)
  y[1] <- y0
  inn <- rnorm(n)
  for (i in 2:n) {
    sds[i] <- sqrt(omega + alpha1 * err[i-1]^2 + beta1 * sds[i-1]^2)
    err[i] <- rnorm(1, mean = 0, sd = sds[i])
    y[i] <- xt * b + err[i]
    # y[i] <- inn[i] * (sqrt(a + b * y[i-1]^2 + c * inn[i-1]^2)) # garch
  }
  y
}
# spec <- garchSpec(model = list())
# y <- garchSim(spec, n = 100)
invisible(capture.output(
  mod_garch <- fGarch::garchFit(formula = ~garch(1, 1), y, include.mean = FALSE)
))
```

```{r}
#| layout: [[100], [50, 50], [50, 50]]
# garch
set.seed(1)
y <- sim_garch(n=500)
plot(y, type = "l")

Acf(y) # supposed to be white noise
Pacf(y)
Acf(y^2) # supposed to be arma pattern
Pacf(y^2) # MA(2) maybe
```


```{r}
#| message: false
#| warning: false
# fitting the garch model
invisible(capture.output(
  mod_garch <- fGarch::garchFit(~garch(1, 1), y,  include.mean = FALSE)
))
summary(mod_garch)
```

+ Jarque-Bera Test - null is normally distributed residuals
+ Ljung-Box - is portmanteau test for autocorrelation, testing for a set of lags.
+ LM arch test - fit the model, and examine the square of the residuals for length of ARCH lags. Null is that there is no AR trend in the squared residuals. See [ARCH wiki](https://en.wikipedia.org/wiki/Autoregressive_conditional_heteroskedasticity)


```{r}
coef(mod_garch) |> kbl() |> kable_minimal(full_width = FALSE)
```



## Vector Autoregression

The simplest vector autoregressive model we can make is VAR(1), which takes the form 

We assume that the process is stable, which implies that the function is stationary.


### Simulation

https://www.r-econometrics.com/timeseries/svarintro/

This is an example of a "B"-model structural, VAR(1) model in which the coefficients are given. Let's see what we can do with given coefficients.

$$
y_t = A_1y_{t-1} + B\epsilon_t
$$
$$
A_1 = \begin{bmatrix} 0.3 & 0.12 & 0.69 \\ 0 & 0.3 & 0.48 \\ 0.24 & 0.24 & 0.3 \end{bmatrix} \text{, }
B = \begin{bmatrix} 1 & 0 & 0 \\ -0.14 & 1 & 0 \\ -0.06 & 0.39 & 1 \end{bmatrix} \text{ and } \epsilon_t \sim N(0, I_3)
$$

```{r}
# Reset random number generator for reproducibility
set.seed(24579)

tt <- 500 # Number of time series observations

# Coefficient matrix
A_1 <- matrix(c(0.3, 0, 0.24,
                0.12, 0.3, 0.24,
                0.69, 0.48, 0.3), 3)

# Structural coefficients
B <- diag(1, 3)
B[lower.tri(B)] <- c(-0.14, -0.06, 0.39)

# Generate series
series <- matrix(rnorm(3, 0, 1), 3, tt + 1) # Raw series with zeros
for (i in 2:(tt + 1)){
  series[, i] <- A_1 %*% series[, i - 1] +  B %*% rnorm(3, 0, 1)
}

series <- ts(t(series)) # Convert to time series object
dimnames(series)[[2]] <- c("S1", "S2", "S3") # Rename variables

# Plot the series
plot.ts(series, main = "Artificial time series")
```

I this case, we've been handed the correct model, and we can quickly note some properties of the model. We can check for stability (which implies stationarity) by looking at the eigenvalues and checking that they're magnitude is less than 1. It's also clear that the innovations are correlated with one another,

```{r}
eigen(A_1)$values
```


```{r}
#| results: hold
var_est <- VAR(series, p = 1, type = "none")

# estimated A_1 matrix
as_latex("\\hat A_1 = ") + as_latex(round(Bcoef(var_est), digits = 3))

# estimate structural equation for A Model
a <- diag(1, 3)
a[lower.tri(a)] <- NA

svar_est_a <- SVAR(var_est, Amat = a, max.iter = 1000)
as_latex("\\hat A^{-1} = ") + as_latex(round(solve(svar_est_a$A), digits = 33)) # pretty close to B

# estimate structural equation for B Model
b <- diag(1, 3)
b[lower.tri(b)] <- NA # specify restriction with NA

svar_est_b <- SVAR(var_est, Bmat = b)
as_latex("\\hat B = ") + as_latex(round(svar_est_b$B, digits = 3))
```

The estimated $A_1$ is given directly by the function `Bcoef` on the fitted VAR model.
We invert the matrix A because we need to translate this into a "B" model from the estimated "A" model, we estimated and find that it's quite correct.

### Moving Average Representations

We can get the basic moving average representation

$$
\begin{aligned}
y_t = \Phi_0 u_t + \Phi_1 u_{t-1} + ...
\end{aligned}
$$

Phi is normally calculated recursively from the last observations, starting with 

1. $\Phi_0 = I_K$
2. $\Phi_i = \sum_{j=1}^{i}\Phi_{i-j}A_j$
  
  + for the VAR(1) case, the sequence is simply $\Phi = I, A_1, A_1^2, A_1^3, \dots$.

```{r}
# Bcoef(var_est) # A_1
# Bcoef(var_est) %^% 2 #A_1^2
Phi(var_est, nstep = 2)
```

For the other MA representation with orthogonal coefficients,

```{r}
# Sigmahat <- crossprod(resid(var_est))/497 # estimated covariance, 500 total obs
# summary(var_est)$covres # i don't know how to get them to match exactly
# P <- t(chol(Sigmahat))
# apply(Phi(var_est, nstep = 2), 3, {\(x) x %*% P}, simplify = FALSE) # multiply all Phi by P
Psi(var_est, nstep = 2)
```

### Impulse Response Analysis

There are two types of basic impulse responses:

1. (forecast error impulse responses) These are on the scale of the original error, `Sigma.u` and let `c(0, 0, 1)` reverberate in the system 

  + these have the disadvantage that
2. (orthogonal error impulse responses) These are on the scale of `Sigma.eps` and let `P^{-1}c(0, 0, 1)` reverberate in the system where $Sigma.u = PP'$ by cholesky decomposition

$$
\begin{aligned}
y_t &= \Phi_0 u_t + \Phi_1 u_{t-1} + \dots \\
&= \underbrace{\Phi_0 P}_{\Psi_0}\underbrace{P^{-1}u_t}_{\epsilon_{t}} +  \dots
\end{aligned}
$$

3. (accumulated error impulse response)


```{r}
#| layout: [[33, 33, 33], [33, 33, 33], [33, 33, 33]]
# First three rows of irf1
# Phi(var_est, nstep = 2) # First columns of this
## even more manually, with A matrix and unit impulse
# Ahat <- Bcoef(var_est)
# Ahat %*% cbind(c(1, 0, 0)) # A e_1
# (Ahat %^% 2) %*% cbind(c(1, 0, 0)) #A_1^2 e_1
irf1 <- irf(var_est, orth = FALSE, boot = FALSE)

# First three rows of irf2
# Psi(var_est, nstep= 2 )
## more manually for VAR(1)
# Sigmahat <- crossprod(resid(var_est))/497 # estimated covariance, 500 total obs
# P <- t(chol(Sigmahat))
# P  %*% cbind(c(1, 0, 0)) # P e_1
# Ahat %*% P %*%cbind(c(1, 0, 0)) # A_1 P e_1
# (Ahat %^% 2) %*% P %*% cbind(c(1, 0, 0)) # A_1 P e_1
irf2 <- irf(var_est, orth = TRUE, boot = FALSE)
# row 2 
# P  %*% cbind(c(1, 0, 0)) + Ahat %*% P %*%cbind(c(1, 0, 0)) 
irf3 <- irf(var_est, cumulative = TRUE, boot = FALSE) # accumulates rowwise of regular irf
plot(irf1)
plot(irf2)
plot(irf3) 
```


The impulse response functions are simply ways of answering the question, "what would happen if I send a shock through the system". For non orthogonal 

We assume the process is mean 0, and $irf_1(0) = c(1, 0, 0)$
$irf_1(1) = A \begin{bmatrix}1, 0, 0\end{bmatrix}$

```{r}
# varsim_orthirf <- irf(var_est, orth = TRUE, boot = FALSE)
# varsim_orthirf$irf$S1
# Bcoef(var_est) # gives as large matrix
```


```{r}
# 
# # ?Psi # MA orthogonal representation
# Psi(var_est) # W = PD^{-1} =
# 
# D <- diag(P)
# W <- P %*% diag((1 / D)) # matches with Psi(1)
# Psi(var_est)[,,1] # I don't know why these are slightly off... but using d
# 
# Psi(var_est)
# 
# P%*% t(P) # Sigmahat

```

```{r}
# P %*% diag(1/D)
# svar_est_b$B
# 
# solve(svar_est_a$A) # This is estimate of B matrix
```


```{r}
# svar_est_a$Sigma.U / 100
# tcrossprod(svar_est_b$B) # is Sigma.U
# tcrossprod(W) # Sigma U = PD^{-1}
# Sigmahat
# solve(W) %*% (svar_est_b$Sigma.U / 100) %*% t(solve(W)) # Identity, Sigma_epsilon
# Psi(var_est)[,,2]
# Acoef(var_est) # gives as list
```


### Example: LA County (VAR)

```{r}
# cardiovascular mortality, temperature and particulates in LA county, weekly.
la <- cbind(cmort, tempr, part)

# visualization
# ts.plot(cmort, tempr, part, col = 1:3) # base r of autoplot 
autoplot(la, color = "black") + facet_grid(series~., scales = "free_y")
```



```{r}
#| render: normal_print
la_var1 <- VAR(la, p = 1, type = "both") # w/ trend
la_var2 <- VAR(la, p = 2, type = "both") # w/ trend

# coefficient matrix
sapply(coef(la_var1),rlang::as_function(~.x[,"Estimate"])) |> as.data.frame()

# vcov of coef estimates
sapply(coef(la_var1), rlang::as_function(~round(.x[,"Std. Error"], 3))) |> as.data.frame()

# estimated covariance of errors
summary(la_var1)$covres |> as.data.frame()
```


```{r}
la_var1_const <- VAR(la, p = 1, type = "const")
# matching manaully with ar.ols
# la_ols <- ar.ols(la, order.max = 1, demean = FALSE, intercept = TRUE)
# 
# # matches constant VAR coefficients
# rbind(t(la_ols$ar[1,,]),
#       const = la_ols$x.intercept)

sapply(coef(la_var1_const),rlang::as_function(~.x[,"Estimate"])) |> as.data.frame()
```


```{r}
la %>% melt(c("time", "series")) %>% 
  ggplot(aes(time, value)) + 
  geom_line(color = "black") + 
  geom_line(data = fitted(la_var1) %>% melt(c("time", "series")),
            mapping = aes(time, value, color = "red"), alpha= .6) +  # var1
  geom_line(data = fitted(la_var1_const) %>% melt(c("time", "series")),
            mapping = aes(time, value, color = "blue"), alpha = .6) + # var1 const
  geom_line(data = fitted(la_var2) %>% melt(c("time", "series")),
            mapping = aes(time, value, color = "green"), alpha = .6) + # var2
  scale_color_manual(values = c("red", "blue", "green"),
                     labels = c("var1", "var1_const", "var2")) + 
  facet_grid(series~.)
```

From looking at the models, they're quite difficult to tell which model is fitting better than the others. 


```{r}
#| include: false
#| eval: false
# var estimates using lm
# Matrix of time series
la_mat <- la %>% `class<-`("matrix")

la_var1_manual <- lm(la_mat ~ 
                       cbind(lag(la_mat), # regress on lag
                             1:nrow(la_mat))) # w/ trend
la_var1_const_manual <- lm(la_mat ~ lag(la_mat)) # const version
```

```{r}
#| include: false
#| eval: false
# coefs match
sapply(coef(la_var1),rlang::as_function(~.x[,"Estimate"])) |> as.data.frame()
coef(la_var1_manual)[c(2:4, 1, 5),] %>% # reorder to match
  `rownames<-`(c("cmort", "tempr", "part", "const", "trend")) |> as.data.frame() 

# var matches
# SSE / n-r-p
# nobs - coefs_estimated - VAR_order
# crossprod(la_var1_const_manual$residuals) / (nrow(la_mat) - la_var1_const_manual$rank - 1) # 503
summary(la_var1_const)$covres |> as.data.frame()
```
We can try to do some automatic VAR order selection with the function `VARselect()`, which uses a number of other criteria.

```{r}
#| echo: true
# order selection
VARselect(la) # selects 2 by BIC
```

The CCF plots should all be non significant. The second part of "x & y" are the ones that lead.

```{r}
#| results: hold
# serial test, BG test for 
serial.test(la_var2, lags.pt = 12, type = "PT.adjusted")
serial.test(la_var2, lags.pt = 12, type = "BG")
serial.test(la_var2, lags.pt = 12, type = "ES")
```

`serial.test` will check for portmanteau, and BG

### sparsevar

Since VAR models can grow $O(np^2)$, this is a high dimensional problem when we're tracking many states. We can use sparsity constraints during fitting.

The theoretical properties of this estimator are studied by VAR

The [SCAD penalty](https://statisticaloddsandends.wordpress.com/2018/07/31/the-scad-penalty/), and [MCP penalty](https://statisticaloddsandends.wordpress.com/2019/12/09/the-minimax-concave-penalty-mcp/) are used in this package

$$
\begin{aligned}
\min_\beta \frac{1}{2} \|y - X\beta\|^2_2 + p(\beta)
\end{aligned}
$$

```{r}
set.seed(1)

# 5% of non-zero entries and a Toeplitz variance-covariance matrix with rho = 0.5.
sim <- simulateVAR(N = 20, p = 2)

fit <- fitVAR(sim$series, p = 2, threshold = TRUE)
plotVAR(sim, fit) # recovered quite well
```
The top image is the true generating values of $A$ in the vector autoregression model.





## State Space Model

These models have a hierarchical form, in which there is some underlying time series process, but then we also observe data on top of that. Thus, the general form of the equations look like this:

$$
\begin{aligned}
\textbf{State Equation:}& \\
x_t &= \Phi x_{t-1} + \Upsilon u_t +   w_t \\
\textbf{Observation Equation:}& \\
y_t &= A_t x_t + \Gamma u_t  + v_t
\end{aligned}
$$
where:

- $\Upsilon u_t$ - is time varying exogenous variables
- $\Upsilon u_t$ - is time varying exogenous variables
- $x_t$ is state at time $t$
- $\Phi x_t$ is state at time $t$ describes how $x$ evolves
- $w_t$ is state noise with $w_t \sim N(0, Q)$
- $v_t$ is measurement noise with $v_t \sim N(0, R)$

There are more general forms of the state space model: with correlated errors,
see Durbin Koopman for more state space methods.

It seems now that state space models are now also being superseded by recurrent
neural networks, which can model dynamical properties.

### Example: AR(1) with observational noise

The state equation:

$$
\begin{aligned}
\textbf{State Equation:}& \\
x_t &= \phi x_{t-1} + w_t \\
\textbf{Observation Equation:}& \\
y_t &= x_t + v_t
\end{aligned}
$$

What is interesting about this case with a hierarchical data structure, is that we can show it has the same error structure as an ARMA model. Shumway Stoffer notes that even though it has the same parameterization as an ARMA model, it is often easier to think about the state model form. See example 6.3 for more details.

## Kalman Filter

Kalman filter is a recursive, markovian updating algorithm for estimating a hidden state variable given noisy and partial observations. The common example is that we are tracking a truck by gps observations. Since the gps observations are imprecise, they will jump back and forth.

An excellent resource explanation with pictures is found [from bzarg](https://www.bzarg.com/p/how-a-kalman-filter-works-in-pictures/)

### Example: Local level Model

Consider the equations:

$$
\begin{aligned}
\textbf{State Equation:}& \\
x_t &= x_{t-1} + w_t \\
\textbf{Observation Equation:}& \\
y_t &= x_t + v_t
\end{aligned}
$$
where both $w_t, v_t \sim N(0, 1)$. We can make the Kalman filter here based on our observation equation. We assume that $A = 1$ (in this case) and $Phi = 1$ are known here. In reality, we can use maximum likelihood of the _innnovations_ (prediction errors) in order to estimate the 

```{r}
# generate the data
set.seed(1)
n <-  50
x0 <- rnorm(1) # random initial state
w <- rnorm(n, 0, 1) # state level noise
v = rnorm(n, 0, 1) # observation level noise
x <- cumsum(c(x0, w)) # true state variables
y <- x[-1] + v # remove initial state
ks <- Ksmooth0(num = 50, 
               y = y, 
               A = 1, 
               mu0 = 10, # set initial values
               Sigma0 = 20, # set initial values
               Phi = 1,
               cQ = 1,
               cR = 1)
```


```{r}
# plot all the predictions
par(mfrow = c(3, 1),
    mar = c(2, 4, 2, 4))
# predictions
plot(x[-1], main = "Prediction", ylim = c(-5, 10))
lines(ks$xp[1,,])
lines(ks$xp + 2*sqrt(ks$Pp[1,,]), lty = 2, col = 4)
lines(ks$xp - 2*sqrt(ks$Pp[1,,]), lty = 2, col = 4) # variance matrices

# filters
plot(x[-1], main = "Filters", ylim = c(-5, 10), xlab = "")
lines(ks$xf[1,,])
lines(ks$xf + 2*sqrt(ks$Pf[1,,]), lty = 2, col = 4)
lines(ks$xf - 2*sqrt(ks$Pf[1,,]), lty = 2, col = 4) # variance matrices


# ks$xs[1,,] # smooth values
plot(x[-1], main = "Smooth", ylim = c(-5, 10), xlab = "")
lines(ks$xs[1,,])
lines(ks$xs + 2*sqrt(ks$Ps[1,,]), lty = 2, col = 4)
lines(ks$xs - 2*sqrt(ks$Ps[1,,]), lty = 2, col = 4) # variance matrices
```

```{r}
# creating table for comparison
tibble(predict = ks$xp[1,,],
       filter = ks$xf[1,,],
       smooth = ks$xs[1,,],
       predict_sd = sqrt(ks$Pp[1,,]),
       filter_sd = sqrt(ks$Pf[1,,]),
       smooth_sd = sqrt(ks$Ps[1,,])) %>%
  mutate(across(predict:smooth, ~scales::number(.x, accuracy = .001)),
         across(predict_sd:smooth_sd, ~scales::number(.x,accuracy = .01))) %>% 
  transmute(predict = glue("{predict} ({predict_sd})"),
         filter = glue("{filter} ({filter_sd})"),
         smooth = glue("{smooth} ({smooth_sd})"))
```


### Estimation of State Parameters

There are bayesian methods for estimating the state parameters but also maximum likelihood w/ Newton Raphson or EM algorithmss

### Example: Nile {.tabset}

Annual flow of the river Nile from 1871 - 1970. There's an apparent changepoint near 1898.

I found this example going through all the state space model libraries. The orginal paper is called "JSS Journal of Statistical Software, State Space Models in R".Specifically, we walk through the Nile example with 3 functions: 

1. `stats::StructTS`
2. `dlm:dlmMLE`
3. `KFAS:kf`

```{r}
# change point analyses
ts.plot(Nile)
```

#### stats::StructTS


```{r}
nile_sts <- StructTS(Nile, "level")
# nile_sts %>% tsdiag() # diagnostics of structural model

# values from model.
tibble(
  times = time(Nile),
  filtered = fitted(nile_sts)[,"level"], # filtered values
  smoothed = tsSmooth(nile_sts)[,"level"]) # smoothed valuees

# forecast values
predict(nile_sts,n.ahead = 5)

# plotting forecast values, with package forecast
plot(forecast::forecast(nile_sts, # StructST object
                   level = c(50, 90), # CI levels
                   h = 10), # periods of forecasting
     xlim = c(1950, 1980))
```

#### dlm::dlm

```{r}
# set up the dlm object
nile_dlm_ll <- function(theta){
  dlmModPoly(order = 1, dV = theta[1], dW = theta[2]) # fits local level model
}

# calls optim internally to optimize model (default BFGS)
# could use numDeriv::hessian for numerically accurate evaluation of Hessians
nile_dlm_mle <- dlmMLE(Nile, # data
                   parm = c(100, 2), # initial parameters
                   nile_dlm_ll, # model
                   lower = rep(1e-4, 2)) 

nile_dlm_mle$par # similar variance parameters of local linear model
```


```{r}
nile_dlm_ll_best <- nile_dlm_ll(nile_dlm_mle$par) # build model with best fit by optim
W(nile_dlm_ll_best) # state randomnesss
V(nile_dlm_ll_best) # observation error variance

# can use fitted model to create smooth estimates now
# $s has time series of smooth estimates
# $U.S and $D.S have SVD of smoothing varriances (for std err)
nile_dlm_ll_smooth <- dlmSmooth(Nile , nile_dlm_ll_best)
# conf ints can be calculated by
```


```{r}
# calculate standard errors
hwidth <- sqrt(unlist(dlmSvd2var(nile_dlm_ll_smooth$U.S, nile_dlm_ll_smooth$D.S))) * qnorm(0.025, lower = FALSE)
nile_dlm_ll_smooth_ci <- cbind(nile_dlm_ll_smooth$s, as.vector(nile_dlm_ll_smooth$s) + hwidth %o% c(-1, 1))
autoplot(nile_dlm_ll_smooth_ci) + theme(legend.position = "") +
  labs(title = "smoothed kalman with CI") +
  geom_point(data = tibble(time = time(Nile),
                           flow = Nile,
                           series = "real"),
             mapping = aes(time, flow))
```

#### KFAS::KFS

Kalman filtering/smoothing/simulation for linear state space models in the exponential family. KFAS uses the sequential processing method. This package uses a slightly different parameterization of their state space models.

$$
\begin{aligned}
y_t &= Z_t\alpha_t + \varepsilon_t &\text{observation}\\
\alpha_{t+1} &= T_t \alpha_t + R_t\eta_t &\text{transition}
\end{aligned}
$$

```{r}
# build the local linear model
# can initialize a specific model with values
# nile_kfas_ll_model <- SSModel(Nile ~ SSMtrend(1, Q = 15000), # Q = Transition error
#                               H = 30) # Observation error

# add NA for values you want to optimize
nile_kfas_ll_model <- SSModel(Nile ~ SSMtrend(1, Q = list(matrix(NA))), # Q = Transition error
                              H = matrix(NA)) # Observation error


# fit the model with wrapper to `optim`
# -logLik(nile_kfas_ll_model) # is the objective that is optimized
nile_kfas_ll_fit <- fitSSM(nile_kfas_ll_model,
       c(log(var(Nile)), log(var(Nile))), # initial parameters for optim
       method = "BFGS")

# extract just the optimal model
nile_kfas_ll <- nile_kfas_ll_fit$model

# Filter/smooth
nile_kfas_ll_smooth <- KFS(nile_kfas_ll,
    filtering = "state",
    smoothing = "state")

# autoplot calls fortify, grabbing some components of the model
# ggfortify:::autoplot.KFS
# fortify(nile_kfas_ll_smooth)  # create df with: time, raw y, alphahat (smoothed values), raw residual = raw y - alphahat
# nile_kfas_ll_smooth$alphahat # contains smoothed state variable estimates. for some reason "fitted" doesn't return these.
autoplot(nile_kfas_ll_smooth)
```


```{r}
# these give same values
# cbind(predict(nile_kfas_ll),
      # nile_kfas_ll_smooth$alphahat)

nile_kfas_ll_ci <- predict(nile_kfas_ll, interval = "confidence", level = .9)
nile_kfas_ll_pi <- predict(nile_kfas_ll, interval = "prediction", level = .9)

                                                                              
legend_labels <- c("lower_ci", "upper_ci", "fit", "lower_pi", "upper_pi")
# forecast:::autoplot.mts
autoplot(cbind(nile_kfas_ll_ci[,-1], nile_kfas_ll_pi),
         mapping = aes(x, y, group = series, linetype = series)) +
  scale_linetype_manual(values = c(2, 2, 1, 3, 3), labels = legend_labels) +
  scale_color_manual(values = c(2, 2, 1, 3, 3), labels = legend_labels) +
  labs(y = "Predicted Annual flow", main = "River Nile")
```


```{r}
# forecasting
nile_kfs_mean <- KFS(nile_kfas_ll,
    filtering = c('state')) # specifying "mean" only estimates the smooths (alphahat)

cbind(predict = nile_kfs_mean$a, # one step ahead prediction
      filter = nile_kfs_mean$att, # filter estimates
      smooth = nile_kfs_mean$alphahat) %>%  # smoothed
  autoplot() + 
  coord_cartesian(ylim = c(700, 1250)) +
  labs(title = "predict/filter/smooth estiamtes of Nile with KFS")
```

