---
title: "Crossover Designs"
date: "`r Sys.Date()`"
author: "Michael Liou"
execute: 
  cache: true
---

```{r setup, include=FALSE}
#| code-summary: Libraries
#| message: false
library(tidyverse)
library(tibble)
library(purrr)
library(effectsize) # for pooled sd
```

## Introduction 

- uniform in sequence
- uniform in period
- balanced
  - carry over effects are all balanced
- strongly balanced
 - including following self

Effects of interest

- "period effect"
- "treatment effect"
- "carryover effect"
- "treatment x period"
- "subject x treatment"
- "subject x period" 

```{r}
tribble(~Advantages, ~Disadvantages,
        "fewer samples", "",
        "more precision per unit", "")
```


## 2x2 Design

### Example: PEF

> The data in Table 3.1 are taken from those which were reported in
Senn and Auclair 1990, p. 1290). They are measurements of peak expiratory
flow PEF), a measure of lung function National Asthma Education Program,
1991, pp. 6±9), made on 13 children aged 7 to 14 with moderate or severe
asthma in a two-treatment two-period cross-over comparing the effects of a
single inhaled dose of 200g salbutamol, a well-established bronchodilator,
and 12g formoterol, a more recently developed bronchodilator Faulds et al.,
1991). The children were randomized to one of two sequence groups. In one
group they were given a dose of formoterol in the morning and observed for 8 h in
the clinic. They then travelled home where they or their parents took further
measurements 10, 11 and 12 h after treatment. PEF is a measurement which
patients may record themselves using a simple device.) On a subsequent occasion
after a wash-out of at least one day they presented at the clinic again and were
given a single dose of salbutamol. Measurements in the clinic followed as before
and were again succeeded by measurements at home. For the second sequence
group, the procedure was as for the first except that they received salbutamol on
the first visit to the clinic and formoterol on the second visit to the clinic.

Peak Exploratory Flow

- `id` - patient id
- `trt` - two levels, formoterol (for) and salbutamol (sal)
- `pef` - peak exploratory flow
- `period` - period of experiment

```{r}
pef <- read.csv("data/pef.csv") %>% 
  # create variable for treatment sequence
  mutate(seq = ifelse((trt == "for" & period == 1) |
                        (trt == "sal" & period == 2),
                      "FS",
                      "SF"),
         trt = dplyr::recode(trt,
                             "for" = "formo",
                             "sal" = "salbu")) %>% 
  rename("trt_seq" = "seq")

pef %>% ggplot(aes(id, pef, color = trt)) +
  geom_point() +
  facet_wrap(~trt_seq)
```

### Analysis

The easiest analysis is probably the paired t-test. This throws away information about the periods in the study.

```{r}
pef_wide <- pef %>% pivot_wider(id_cols = c(id, trt_seq), names_from = trt, values_from = "pef") %>% 
  mutate(diff = formo - salbu)
pef_wide$diff %>% mean() # 45.385
pef_wide$diff %>% sd() # 40.593
pef_t <- t.test(pef_wide$formo, pef_wide$salbu, paired = TRUE)

45.384462 / 4.0312 # the estimated standard error

pef_t
```

The next analysis is to adjust for the period effect with just to do a sample t test of the differences by period.

```{r}

pef_FS <- pef_wide %>% filter(trt_seq == "FS") %>% pull(diff)
pef_SF <- pef_wide %>% filter(trt_seq == "SF") %>% pull(diff)

pef_FS_SS <- sum((pef_FS - mean(pef_FS))^2) # FS sum squares
pef_SF_SS <- sum((pef_SF - mean(pef_SF))^2) # SF sum squares

((length(pef_FS) - 1) * pef_FS_SS + (length(pef_SF) -1) * pef_SF_SS) / (length(pef_FS) + length(pef_SF) - 2)

(pef_FS_SS + pef_SF_SS) / (length(pef_FS) + length(pef_SF) - 2) # 1500.81

sd_pooled(pef_FS, pef_SF)

# pooled estimate of variance

t.test(scale(pef_FS, scale = F), scale(pef_SF, scale = F), var.equal = FALSE)
```



